---
title: "Data 608 - Module 1 Assignment"
output:
  html_document: default
  html_notebook: default
---

**Principles of Data Visualization and Introduction to ggplot2**

I have provided you with data about the 5,000 fastest growing companies in the US, as compiled by Inc. magazine. lets read this in:

```{r}

inc <- read.csv("https://raw.githubusercontent.com/charleyferrari/CUNY_DATA_608/master/module1/Data/inc5000_data.csv", header= TRUE)

```

And lets preview this data:

```{r}

head(inc)

```
```{r}

summary(inc)

```


```{r}
str(inc)
```

Think a bit on what these summaries mean. Use the space below to add some more relevant non-visual exploratory information you think helps you understand this data:

```{r warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(kableExtra)
```

### Growth Rate 

I noticed that the growth rate goes from 0.340 to 421.480. Below, you will see that there are 19 companies that experienced growth rates of 100 or higher. 

```{r}
inc %>% dplyr::filter(Growth_Rate >= 100) %>% summarise(n = n())
```

Below is the list of these 19 companies with growth rates of 100 or higher. 

```{r}
kable(inc %>% dplyr::filter(Growth_Rate >= 100)) %>% kable_styling()
```

### Revenue

The revenue ranges from 2 million to about 10 billion. The median revenue is about 11 million.  

```{r}
inc %>% dplyr::summarise(min=min(Revenue), median=median(Revenue), max=max(Revenue))
```

### Industry

There are 25 distinct industries. 

```{r}
kable(inc %>% dplyr::group_by(Industry) %>% dplyr::summarise(n=n()) %>% arrange(desc(n))) %>% kable_styling()
```

### Employees

There are some companies that do not have data for Employee. The number of employees range from 1 to 66,803. The median employee size is 53. 

```{r}
kable(inc %>% dplyr::summarise(min=min(Employees, na.rm = TRUE), median=median(Employees, na.rm = TRUE), max=max(Employees, na.rm = TRUE))) %>% kable_styling()
```

### City

There are 1,519 distinct cities. 

```{r}
cities <- inc %>% group_by(City) %>% summarise(n=n())
nrow(cities)
```

These are the top 10 cities (based on the number of companies that are located in the city). 

```{r}
kable(inc %>% group_by(City) %>% summarise(n=n()) %>% arrange(desc(n)) %>% top_n(10)) %>% kable_styling()
```

### State

There are 52 distinct states in the data set. 

```{r}
states <- inc %>% group_by(State) %>% summarise(n=n())
nrow(states)
```

These are the top 10 States (based on the number of companies that are located in the State). 

```{r}
kable(inc %>% group_by(State) %>% summarise(n=n()) %>% arrange(desc(n)) %>% top_n(10)) %>% kable_styling()
```

---

## Question 1

Create a graph that shows the distribution of companies in the data set by State (ie how many are in each state). There are a lot of States, so consider which axis you should use. This visualization is ultimately going to be consumed on a 'portrait' oriented screen (ie taller than wide), which should further guide your layout choices.


```{r}

# Answer Question 1 here

ordered <- inc %>% group_by(State) %>% summarise(n=n()) %>% arrange(desc(n))

plt1 <- 
  ggplot(data = ordered[1:52,], aes(x=reorder(State,n), y=n)) + 
  geom_bar(stat="identity", width=0.5, color="#1F3552", fill="steelblue", 
           position=position_dodge()) +
    #geom_text(aes(label=round(n, digits=2)), hjust=1.3, size=3.0, color="white") + 
    coord_flip() + 
    scale_y_continuous(breaks=seq(0,700,100)) + 
    ggtitle("Disbribution by State") +
    xlab("") + ylab("") + 
    theme_minimal()
```

I couldn't find a way to increase the plot canvas size. This would look better if there's more space in between each state, and the bars are slightly bigger. 

The graph below orders the distribution from highest to lowest states. 

```{r}
plt1
```


---


## Quesiton 2

Lets dig in on the state with the 3rd most companies in the data set. Imagine you work for the state and are interested in how many people are employed by companies in different industries. Create a plot that shows the average and/or median employment by industry for companies in this state (only use cases with full data, use R's `complete.cases()` function.) In addition to this, your graph should show how variable the ranges are, and you should deal with outliers.

As you can see, the state with the 3rd most companies in the data set is New York. 

```{r}
kable(inc %>% group_by(State) %>% summarise(n=n()) %>% arrange(desc(n)) %>% top_n(3)) %>% kable_styling()
```


`inc_cc` holds complete cases only. 

```{r}
inc_cc <- inc[complete.cases(inc),]
```

Below is a break down of median number of employees in each industry for New York state. It shows the min, median, and max number of employees for each industry in NY. It is ordered from highest to lowest variability. 

```{r}
kable(inc_cc %>% filter(State=='NY') %>% group_by(Industry) %>% summarise(min=min(Employees),median=median(Employees), max=max(Employees), var=var(Employees)) %>% arrange(desc(var))) %>% kable_styling()
```

A box plot could show the median number employees (this is indicated by the dark black line in the box). A box plot also shows the range of the data and outliers (indicated by a red asterisk symbol). 

There are 25 different industries. I tried plotting them all in a single box plot call, and the result was too tiny to get any kind of useful information. In question 1, I also had a similar problem of properly spacing out the data elements on the screen. As a workaround, I created vectors that group records based on variability. The table above was used for this purpose. In this case, companies that have higher variability in number of employees are also ones with higher maximum number of employees. 

The code below groups industries together with similar variability. I try to limit each group up to 5 industries so that the plot doesn't get too small. 


```{r}
g1a <- c('Business Products & Services')
g1b <- c('Consumer Products & Services')
g2 <- c('Travel & Hospitality', 'Human Resources', 'IT Services', 'Software')
g3 <- c('Security', 'Media', 'Financial Services',  'Environmental Services', 'Food & Beverage')
g4 <- c('Energy', 'Telecommunications', 'Manufacturing', 'Health', 'Construction')
g5 <- c('Advertising & Marketing', 'Education', 'Engineering', 'Logistics & Transportation', 'Retail')
g6 <- c('Insurance', 'Real Estate', 'Computer Hardware', 'Government Services')
```


Below is the code for creating the box plots for each grouping. 

Please note that each plot for each group has a different x-axis scale, which depends on the range of number of employees for each respective group. 

The industries 'Computer Hardware' and 'Government Services' do not have enough data to generate a box plot. 

```{r}
plt_g1a <- ggplot(inc_cc %>% filter(State=='NY' & Industry %in% g1a), aes(x = Industry, y = Employees)) + 
        coord_flip() + 
        geom_boxplot(outlier.colour="red", outlier.shape=8,
             outlier.size=1, notch=FALSE)

plt_g1b <- ggplot(inc_cc %>% filter(State=='NY' & Industry %in% g1b), aes(x = Industry, y = Employees)) + 
        coord_flip() + 
        geom_boxplot(outlier.colour="red", outlier.shape=8,
             outlier.size=1, notch=FALSE)

plt_g2 <- ggplot(inc_cc %>% filter(State=='NY' & Industry %in% g2), aes(x = Industry, y = Employees)) + 
        coord_flip() + 
        geom_boxplot(outlier.colour="red", outlier.shape=8,
             outlier.size=1, notch=FALSE)

plt_g3 <- ggplot(inc_cc %>% filter(State=='NY' & Industry %in% g3), aes(x = Industry, y = Employees)) + 
        coord_flip() + 
        geom_boxplot(outlier.colour="red", outlier.shape=8,
             outlier.size=1, notch=FALSE)

plt_g4 <- ggplot(inc_cc %>% filter(State=='NY' & Industry %in% g4), aes(x = Industry, y = Employees)) + 
        coord_flip() + 
        geom_boxplot(outlier.colour="red", outlier.shape=8,
             outlier.size=1, notch=FALSE)

plt_g5 <- ggplot(inc_cc %>% filter(State=='NY' & Industry %in% g5), aes(x = Industry, y = Employees)) + 
        coord_flip() + 
        geom_boxplot(outlier.colour="red", outlier.shape=8,
             outlier.size=1, notch=FALSE) 

plt_g6 <- ggplot(inc_cc %>% filter(State=='NY' & Industry %in% g6), aes(x = Industry, y = Employees)) + 
        coord_flip() + 
        geom_boxplot(outlier.colour="red", outlier.shape=8,
             outlier.size=1, notch=FALSE)
```


I created a separate group for 'Business Products & Services' and 'Consumer Products & Services' because the box plots for these came out so tiny. It looks like the outlier data is causing the box plot of these 2 industries to flatten out too much. 


```{r}
plt_g1a
plt_g1b
```

Below are the box plots for the rest of the other industries. 

Please be mindful that the x-axis scale for each grouping is different. 

```{r echo=FALSE}
plt_g2
plt_g3
plt_g4
plt_g5
plt_g6
```

---

## Question 3

Now imagine you work for an investor and want to see which industries generate the most revenue per employee. Create a chart that makes this information clear. Once again, the distribution per industry should be shown.

The table below shows the total number of companies in each industry and the revenue per employee for each industry. 

```{r}
revenue_per_employee <- 
inc_cc %>% group_by(Industry) %>% summarise(count=n(), total_revenue=sum(Revenue), total_employees=sum(Employees), revenue_per_employee=total_revenue/total_employees) %>% arrange(desc(revenue_per_employee))

kable(revenue_per_employee) %>% kable_styling()
```


The code below plots the revenue per employee as a bar chart sorted by revenue per employee from highest to lowest. A second bar chart plot is generated that shows the distribution of companies by industry sorted by revenue per employee from highest to lowest (same order as the first plot). 


```{r}
plt3_1 <- ggplot(data=revenue_per_employee, aes(x=reorder(Industry,-revenue_per_employee), y=revenue_per_employee)) +
     geom_bar(stat="identity", fill="steelblue") +
     theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
     ggtitle("Revenue Per Employee by Industry") + 
     ylab("Revenue Per Employee") + 
     xlab("")

plt3_2 <- ggplot(data=revenue_per_employee, aes(x=reorder(Industry,-revenue_per_employee), y=count)) +
     geom_bar(stat="identity", fill="steelblue") +
     theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
     ggtitle("Distribution of Companies by Industry") + 
     ylab("Revenue Per Employee") + 
     xlab("")
```



```{r echo=FALSE}
plt3_1
plt3_2
```

